Simple Type Theory is not too Simple

### 2020-12-23 article.tex
improved introduction

It is well-known in the Isabelle community that the algebra library should be rebuilt from scratch. First, I should say the algebra libraries. Indeed, there are many libraries for algebra in Isabelle, the \textit{HOL-Algebra}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Algebra/index.html}} started in the previous millennium (1999), the more recent \textit{HOL-Computational\_Algebra}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Computational_Algebra/index.html}}, the oldest formalization of algebraic structures (groups, rings, modules, fields, vector spaces \dots) being found in the main library \textit{HOL}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL/index.html}}, and this is in addition to some entries in the Archive of Formal Proofs\footnote{https://www.isa-afp.org/} like \textit{Groups, Rings and Modules} \cite{Group-Ring-Module-AFP} and \textit{Vector Spaces} \cite{VectorSpace-AFP} to cite a few. As a result, there is an important amount of duplicated code and it makes the situation unclear regarding the status of algebra for the newcomer in the Isabelle world. Second, if these formalizations constitute an impressive body of work, not without its merits, parts of it are deprecated code not using the Isar language or the powerful locales, a module system in Isabelle for managing theory hierarchies \cite{ballarin2014}. A successful experiment was conducted recently in \cite{ballarin20} showing that locales allow for a smooth handling of basic algebraic structures in Isabelle. So, hopefully, the situation should improved in the near future. \\

@@

In this article we present a concise and elegant formalization of schemes in Isabelle/HOL following the classic graduate textbook of Hartshorne \cite{hartshorne}.

@@

We do not use typeclasses, we do use locales. We encapsulated within the locale dedicated to a preasheaf of rings $\mathscr{F}$ the ring structures on the underlying sets $\mathscr{F}(U)$'s simply using the functional \textit{add\_str} (\textit{resp.} \textit{mult\_str}, \textit{zero\_str}, \textit{one\_str}) that takes an open subset $U$ and outputs the multiplication (\textit{resp.} the addition, the additive unit, the multiplicative unit) on the set $\mathscr{F}(U)$. The resulting formalization of sheaves of rings in Isabelle is not only possible, but it is concise (6 lines), very neat and tidy, almost transparent with respect to the pen-and-paper definition. \\
It will not be any more difficult to define (pre)sheaves valued in any category of structured sets (groups, modules \dots) or even to define in Isabelle a general notion of $\mathscr{C}$-valued (pre)sheaves by embarking the category $\mathscr{C}$ as part of the data within the locale.

>>>>

It is well-known in the Isabelle community that the algebra library should be rebuilt from scratch. First, I should say the algebra libraries. Indeed, there are many libraries for algebra in Isabelle, the \textit{HOL-Algebra}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Algebra/index.html}} started in the previous millennium (1999), the more recent \textit{HOL-Computational\_Algebra}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Computational_Algebra/index.html}}, the oldest formalization of algebraic structures (groups, rings, modules, fields, vector spaces \dots) being found in the main library \textit{HOL}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL/index.html}}, and this is in addition to some entries in the Archive of Formal Proofs\footnote{https://www.isa-afp.org/} like \textit{Groups, Rings and Modules} \cite{Group-Ring-Module-AFP} and \textit{Vector Spaces} \cite{VectorSpace-AFP} to cite a few. As a result, there is an important amount of duplicated code and it makes the situation unclear regarding the status of algebra for the newcomer in the Isabelle world. Second, if these formalizations constitute an impressive body of work, not without its merits, parts of it are deprecated code not using the Isar language or the modern locales, a powerful module system in Isabelle for managing theory hierarchies \cite{ballarin2014}. Therefore, HOL-Algebra contains some peculiarities, a notable one being its locale \emph{abelian\_group} (in \textit{Theory Ring}) which presents the odd twist of requiring a ring structure. Indeed, this last locale uses records and there does not exist multiple inheritances for records, hence the awkward use of the ring record for abelian groups. However, a successful experiment was conducted recently in \cite{ballarin20} showing that modern locales, without records, allow for a smooth handling of basic algebraic structures in Isabelle. So, hopefully, the situation should dramatically improved in the near future. \\

@@

In this article we present a concise and elegant formalization of schemes in Isabelle/HOL based on state-of-the-art locales, without using any record. Our presentation follows the classic graduate textbook of Hartshorne \cite{hartshorne}.

@@

We do not use old-fashioned typeclasses, we do use modern locales. We encapsulated within the locale dedicated to a preasheaf of rings $\mathscr{F}$ the ring structures on the underlying sets $\mathscr{F}(U)$'s simply using the functional \textit{add\_str} (\textit{resp.} \textit{mult\_str}, \textit{zero\_str}, \textit{one\_str}) that takes an open subset $U$ and outputs the multiplication (\textit{resp.} the addition, the additive unit, the multiplicative unit) on the set $\mathscr{F}(U)$. The resulting formalization of sheaves of rings in Isabelle is not only possible, but it is concise (6 lines), very neat and tidy, almost transparent with respect to the pen-and-paper definition. \\
It will not be any more difficult to define (pre)sheaves valued in any category of structured sets (groups, modules \dots) or even to define in Isabelle a general notion of $\mathscr{C}$-valued (pre)sheaves by embarking the category $\mathscr{C}$ as part of the data within the locale.
